---
title: "Constrained analyses"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(vegan)
library(ggvegan)
data(dune)
data(dune.env)
data(varespec)
data(varechem)
knitr::opts_chunk$set(echo = FALSE)
```


## Multiple responses plus explanatory variables
### Introduction
Unconstrained ordination methods such as PCA, CA and NMDS allow you to summarise
the relationships between your samples (sites, isolates, quadrats etc.) and your
attributes (species, gene sequences etc.). They provide a useful method to 
simplify your data so that it can be viewed in a 2-dimensional ordination plot.
The scores from these plots, especially the first axis, can sometimes be related
to potential explanatory variables to aid interpretation, as you showed with
soil moisture and the dune vegetation analysis.

If you have explanatory variables, you might be tempted to extract an ordination
axis, and after visualising any patterns with a potential explanatory variable,
undertake a linear model, with your chosen **ordination axis** as your response.
For example, recall that you undertook a PCA of the sand dune vegetation data,
and showed a clear pattern with Moisture:

```{r dune_pca, echo=TRUE}
# PCA of dune data
dune_pca <- rda(dune)

# Plot of PC1 vs PC2
autoplot(dune_pca, layers="sites", legend.position="none", geom="text",
         title = "PCA of dune vegetation; site codes.")

# Extract PC1 and relate to soil moisture category
dune_pc1 <- scores(dune_pca, display="sites", choices = 1)
gf_boxplot(dune_pc1 ~ Moisture, data=dune.env)
```

Now let's do a linear model to formally test the relationship between soil
moisture and dune vegetation composition as described by PC1. We'll display
the ANOVA table, and check the first two model diagnostic plots (residuals and
QQ plots):

```{r dune_pca_lm, echo=TRUE}
dune_lm <- lm(dune_pc1 ~ Moisture, data=dune.env)
anova(dune_lm)
mplot(dune_lm, which=1:2)

```

The relationship with moisture class is highly significant, with F=14.94426 and
p=6.691419e-05 which is p=0.00006691419 (**remember** you would report these as
"F=14.94, p<0.001"). There are no obvious problems with the residuals vs fitted
plot, with an even scatter around the zero line. The QQ plot looks good, with
most points along the expected diagonal line.

When techniques to handle lots of response variables were first developed, this
was the most common method of analysis. It is sometimes referred to as **indirect gradient analysis**
and was widely used until the 1990's.

## Problems with indirect gradient analysis
The linear model presented on the previous page showed no obvious problems, so
the disadvantages of indirect gradient analysis may not be immediately obvious.
However, one assumption of linear models (and GLMs), is that all your response
data points are independent of each other:

* The composition of plants in your first quadrat should not affect those in your second. 
* The gene sequence from your third isolate should not change those in your fifth
* The bacterial OTU samples from Chile should be independent of the OTUs obtained from France
* The types of insects found in your pitfall trap sample from Northumberland should
not influence those found in your pitfall trap from Cornwall

This seems fairly obvious, and in practical terms, when you collect the data from
field surveys or laboratory experiments, the various samples are independent. The
problem arises from what happens when you undertake an unconstrained ordination.

### Non-independence of ordination scores
Let's repeat our PCA of the sand dune vegetation, but omit one of the samples
at random (sample 6):

```{r compare_pca}
# Multiple plot function from http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/ 
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

dune_pca <- rda(dune)
dune_pca_no6 <- rda(dune[-6,])
dune_plt1 <- autoplot(dune_pca, display="sites", geom="text", legend.position = "none",
                      title="PCA with full dataset")
dune_plt2 <- autoplot(dune_pca_no6, display="sites", geom="text", legend.position = "none",
                      title="PCA with one sample omitted")
multiplot(dune_plt1, dune_plt2, cols=2)
```

You can see that the whole ordination has 'flipped' on its vertical axis. Depending
on your data, sometimes both PC1 and PC2 will flip or rotate. The relative positions
of the samples are still roughly the same, in the samples similar in their species
composition (e.g. 17, 18 and 19) are still relatively close to each other, but they
have nevertheless moved.

**Key point**
Whilst your original quadrats, isolates, samples or sites may have been independent
from each other, once they are converted to PCA axis scores, the actual scores are
**not** independent. Fortunately, an alternative method to resolve this problem,
known as **constrained ordination** was developed in 1989, and has since become
a standard technique for biologists.
