---
title: "BIO2020 - Constrained ordination"
author: "Roy Sanderson"
date: "07/12/2020"
output: html_document
---

```{r setup, include=FALSE}
library(bio2020)
data("varespec")
data("varechem")
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Unconstrained ordination methods such as PCA, CA and NMDS allow you to summarise
the relationships between your samples (sites, isolates, quadrats etc.) and your
attributes (species, gene sequences etc.). They provide a useful method to 
simplify your data so that it can be viewed in a 2-dimensional ordination plot.
The scores from these plots, especially the first axis, can sometimes be related
to potential explanatory variables to aid interpretation, as you showed with
soil moisture and the dune vegetation analysis.

If you have explanatory variables, you might be tempted to extract an ordination
axis, and after visualising any patterns with a potential explanatory variable,
undertake a linear model, with your chosen **ordination axis** as your response. This is sometimes also called **indirect gradient analysis**. However, one assumption of linear models (and GLMs), is that all your response data points are independent of each other:

* The composition of plants in your first quadrat should not affect those in your second. 
* The gene sequence from your third isolate should not change those in your fifth
* The bacterial OTU samples from Chile should be independent of the OTUs obtained from France
* The types of insects found in your pitfall trap sample from Northumberland should
not influence those found in your pitfall trap from Cornwall

This seems fairly obvious, and in practical terms, when you collect the data from
field surveys or laboratory experiments, the various samples are independent. The
problem arises from what happens when you undertake an unconstrained ordination.

The challenge is that the sample scores within an ordination axis are **not** independent of each other. Indeed if a single sample is removed from your dataset
there is the risk that all the others will shift, although their **relative** positions to each other will remain constant. See [this example of the problems in a PCA](https://naturalandenvironmentalscience.shinyapps.io/Constrained/#section-problems-with-indirect-gradient-analysis) when one row is removed. One assumption of conventional linear models is that each sample (row) is independent, yet it is obvious that ordination scores breach this assumption.

Fortunately, an alternative method to resolve this problem,
known as **constrained ordination** was developed in 1989, and has since become
a standard technique for biologists.

## Constrained ordination
In a constrained ordination the explanatory variables (categorical and/or
continuous) are incorporated into the ordination itself. The sample scores are
constrained to be linear combinations of the various explanatory variables,
whilst simultaneously accounting for the composition of the attributes. So the
overall format is:

$$\textit{Table of response variables} = \textit{Explanatory variables} + \epsilon$$

Note that the technique does not work effectively if you only have one or two
explanatory variables, as it may constrain all your samples or attributes too
much along one axis. The display of constrained analysis is in the form of 
modified ordination plots, which can be very informative once you have learnt
how to interpret them. You can also undertake formal statistical tests using
analyses analogous to ANOVA. The technique will also cope with complex experimental
designs, such as blocked designs, or time-series. You can also create interaction
terms if needed. Constrained analysis exists in two main forms, linear and unimodal. The linear form is Redundancy Analysis (RDA) and unimodal is Canonical Correspondence Analysis (CCA). These are run using the `ordi_rda()` and `ordi_cca()` functions respectively.

The overall aim of this practical is to introduce you to the analysis, interpretation and display of multi-variable data with explanatory variables through constrained ordination. Specific objectives are to:

1. Use unimodal methods to analyse, visualise and interpret the ordination plots
2. Demonstrate "permutation ANOVA" of the whole analysis, different explanatory variables, and individual ordination axes
3. Use stepwise selection to identify the minimum number of explanatory variables.

The approach for the analysis and interpretation of linear methods is very similar.

# 1. Unimodal analysis and visualisation
## 1.2 Study dataset
The study dataset is one we have looked at already, that comes with the `vegan` package, namely the `varespec` dataset reindeer-grazed lichen pasture data published in Vare et al (1995). This has 24 rows (samples) and 44 columns (species) and form our "response" dataset. There is also a `varechem` dataset, with 14 columns containing soil characteristics. These are the "explanatory" dataset, and of course 14 explanatory variables is a very large number and difficult to interpret. This is where constrained ordination can help us simplify the data to aid understanding. The basic model we therefore have is:


$$\textit{Tundra vegetation} = \textit{Soil chemistry} + \epsilon$$

This is a vegetation ecology example. However, I have published papers using this method on gene sequences from microbial data, and it can be applied to any discipline where you have large numbers of potential response and explanatory variables.

Begin by loading the two datasets, and explore their contents, for example

```{r, eval=FALSE}
library(bio2020) # auto-loads the vegan library

# Access the example datasets
data("varespec")
data("varechem")

# Explore the data
summary(varespec)
summary(varechem)
View(varespec)
View(varechem)
```


